---
title: "ARIMA Revenue Forecasting"
author: "Tom Trinter"
date: "December 9, 2019"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(gridExtra)
library(scales)
library(xlsx)
library(data.table)
library(fpp2)
library(fUnitRoots)
library(dplyr)

```

# Revenue Forecasting - ARIMA Modeling

## Objective
The objective of this is to model product revenues to better inform their forecasting process. This is derived from a Fortune 500 client project where we had three years of monthly actual revenue data across 45 different products. These produts are grouped into 6 product groups. The data has been sanitized and product names genericized to protect the proprietary nature of the data and client privacy.

We'll begin by modeling the revenues at the product-group level using simple ARIMA models. We'll then look at modeling at the individual product level. 

## Data

For this analysis, we're starting with a spreadsheet where the data was maintained manually for years in the treasury department. I'll convert the data from its 'wide' form into 'long' for easier use going forward.

#### Documentation

This approach follows the example found [here](https://datascienceplus.com/time-series-analysis-using-arima-model-in-r/).

```{r data}
working.dir <- '/Users/tttri/git_repo/RevenueForecasting/'
setwd(working.dir)

long.filename <- 'Revenue Data - Long.xlsx'
pd <- read.xlsx(long.filename, sheetName = 'forecast_data')

# Reduce to actuals only (forecast_id=0)
pd <- pd %>% 
  subset(forecast_id==0 & month_start>='2017-04-01') %>% 
  mutate(value_m = value/1000000)

```

## Plot Product Groups

```{r Product_Groups}

# Let's plot each product_group vs. time
plot.data <- pd %>%
  select(summary_display, month_start, value_m) %>% 
  group_by(summary_display, month_start) %>%
  summarize(group_value = sum(value_m,na.rm=T)) %>% 
  filter(!is.na(summary_display) & month_start < '2019-07-01')

p <- 
  ggplot(plot.data,
         aes(x=month_start, y=group_value)) + 
  geom_line() + 
  geom_point() +
  geom_smooth(se=F)+
  labs(x="Month",
       y="Value ($MM)",
       title="Monthly Revenue")+
  theme_bw() + 
  facet_wrap(~summary_display, scales="free") 
p
```

## ARIMA Modeling

Looking at the time-series plots of grouped product revenues, there seem to be some potential trends to extract before we can use an ARIMA model. Let's take a closer look at the Annuity data and add in a 1st difference to see if that will get closer to stationary.

```{r Brass_Data}
brass.data <- pd %>%
  filter(summary_display=='Brass' & month_start < '2022-06-01' ) %>% 
  group_by(month_start) %>%
  summarize(group_value = sum(value_m,na.rm=T)) %>% 
  select(month_start, group_value)

ts.data <- ts(brass.data$group_value, frequency=12)

ggtsdisplay(ts.data, main="Brass Revenue")

```

## Components

```{r Components}
# components.ts <- decompose(ts.data)
components.ts <- stl(ts.data, s.window=12)
plot(components.ts)

```

## Adjust for seasonality

From the decomposition plot above, it looks like there is some quarterly seasonality as well as an overall decreasing trend. 

So - we can remove both of those and fit a regression to them separately? For now, just removing the trend...
```{r}
ts.data.adj <- ts.data - components.ts$time.series[,2]
```

## Unit Root Test:

```{r unitRoots}
urkpssTest(ts.data.adj, type=c("tau"), lags = c("short"), use.lag=NULL, doplot=TRUE)

tsstationary <- diff(ts.data.adj, differences=1)
plot(tsstationary)

```

```{r Arima}
fit <- auto.arima(ts.data.adj, trace = TRUE)

```
```{r}
checkresiduals(fit)
```

```{r}
brass.forecast <- forecast(fit)
autoplot(brass.forecast)
```

## Model the trend component
[Example LM](https://datascienceplus.com/fitting-polynomial-regression-r/)
```{r}
 trend.data <- data.frame('days' = brass.data$month_start-brass.data$month_start[1],
                         trend_val = components.ts$time.series[,2] ) %>% 
  mutate(days = as.numeric(days))

simple.fit <- lm(trend_val ~ poly(days, 4), data= trend.data)

summary(simple.fit)

trend.data$predicted <- simple.fit$fitted.values

ggplot(trend.data, aes(days, y = trend_val, color = variable)) + 
    geom_line(aes(y = trend_val, col = "Actual"), size=1) + 
    geom_line(aes(y = predicted, col = "Predicted"), size=1)

```

## Final Forecast

```{r}
brass.data$predicted <-simple.fit$fitted.values + brass.forecast$model$fitted

ggplot(brass.data, aes(month_start, y = group_value, color = variable)) + 
    geom_line(aes(y = group_value, col = "Actual"), size=1) + 
    geom_line(aes(y = predicted, col = "Predicted"), size=1)

```
